---
title: "DATS 6101 Project 2"
author: "Syed Ibrahim Hamza, Pratik Shetty, and Dominique Howell"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: united
    code_folding: hide
---

```{css, echo=FALSE}
/* Font adjustments */
body { font-size: 14px; line-height: 1.5; }
h1 { font-size: 26px; }
h2 { font-size: 20px; }
h3 { font-size: 18px; }
.tocify-content { font-size: 14px !important; }
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(lattice)
library(corrplot)
library(car)
library(caret)
library(leaps)
library(pROC)
library(pscl)
library(rpart)
library(rpart.plot)
library(randomForest)
library(glmnet)
library(MASS)
library(ezids)

brfss = readRDS("../Data/Processed/brfss_2018_2023.rds")
```

# Introduction

## Project Overview
This project examines the **impact of COVID-19 on obesity and lifestyle behaviors** among U.S. adults using the **CDC Behavioral Risk Factor Surveillance System (BRFSS)** data from **2018–2023**.  
The BRFSS is a nationally representative health survey that collects information on demographics, chronic health conditions, and behavioral risk factors from approximately **400,000 adults annually** across all 50 states, Washington D.C., and U.S. territories.  
Our primary goal is to evaluate how obesity prevalence and lifestyle behaviors such as physical activity and alcohol use have changed before and after the onset of COVID-19, and how these trends vary by demographic and health factors.


## Data Sources
- **Raw data**: stored in `Data/Raw/`, obtained from the [CDC BRFSS Annual Data Portal](https://www.cdc.gov/brfss/annual_data/annual_data.htm).  
- **Processed data**: cleaned and merged dataset saved in `Data/Processed/`.  
- **Final analytical dataset**: `Data/Processed/brfss_2018_2023.rds` (over 100k observations, ~24 variables).  

# Question # 1
<br>
Obesity is a major public health concern in the United States and is associated with numerous chronic health conditions.
Using Behavioral Risk Factor Surveillance System (BRFSS) data from 2020–2023, this study examines whether demographic and lifestyle characteristics can be used to predict adult obesity status.

Research Question:

Can we predict the likelihood of an adult being obese (BMI ≥ 30) based on demographic and lifestyle factors such as age, sex, race/ethnicity, income category, physical activity, and alcohol consumption?
<br>

## Data Preparation & Variable Selection
<br>
Body Mass Index (BMI) was used to define obesity status following CDC guidelines. A binary obesity indicator was created where individuals with a BMI of 30 or higher were classified as obese, and those with a BMI below 30 were classified as non-obese. This outcome variable was represented both as a numeric indicator (0/1) for modeling purposes and as a factor variable to support classification-based methods.
<br>

```{r}
brfss = readRDS('../Data/Processed/brfss_2018_2023.rds')

brfss_postcovid <- subset(brfss, interview_year >= 2020)

brfss_postcovid <- brfss_postcovid %>%
  mutate(
    obese_num = ifelse(bmi >= 30, 1, 0),
    obese = factor(obese_num, labels = c("No", "Yes"))
  )
```

<br>
Variables were selected based on prior public health literature and their theoretical relevance to obesity risk. The final analytic dataset includes demographic characteristics, lifestyle behaviors, and temporal indicators that may influence obesity outcomes.
<br>

```{r}
vars <- brfss_postcovid %>%
  dplyr::select(obese, obese_num, bmi, age, sex, race_ethnicity, income_category,
                any_exercise_last_month, binge_drinking,
                max_drinks_30day, interview_year)

glimpse(vars)
```











